---
title: "Minimal_code_example_CP1"
author: "Testing purposes."
date: "`r Sys.Date()`"
output: html_document
---

# Load R libraries

These libraries will be used in this computer session. Note: tidyverse and DESeq2 both have a `select` function. Because we load DESeq2 after tidyverse, you will have to refer to the dplyr select function as `dplyr::select`.

We also strongly advice that you clean up your environment on the right top with the broom symbol before you start today, to remove data from previous computer sessions.

```{r}
library(tidyverse)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("DESeq2", quietly = TRUE))
  BiocManager::install("DESeq2")
library(DESeq2)
```

# Create fake dataset

```{r}
# Parameters
num_genes <- 1000       # Number of genes
num_samples <- 10        # Number of samples
prob <- 0.01              # Probability for the binomial distribution

# Simulating sample conditions (e.g., control vs treatment)
conditions <- factor(rep(c("control", "treatment"), each = num_samples / 2))

# Simulating count data
set.seed(123)  # For reproducibility
counts <- matrix(rbinom(num_genes * num_samples, size = 1000, prob = prob),
                 nrow = num_genes, ncol = num_samples)
counts[1:50,1:5] <- counts[1:50,1:5] + 20 # first 50 control are a bit differentially expressed
counts[51:100,1:5] <- counts[51:100,1:5] + 50 # second 50 control are more differentially expressed
counts[901:1000,] <- 0 # Last 100 are not expressed

colnames(counts) <- paste0("Sample", 1:num_samples)
rownames(counts) <- paste0("Gene", 1:num_genes)
```

```{r}
counts %>%
  data.frame() %>%
  rownames_to_column("SYMBOL") %>%
  pivot_longer(!SYMBOL, names_to = "sample", values_to = "count") %>% 
  ggplot(aes(x = log2(count+1), color = sample)) +
    geom_density() +
    theme(legend.position = "bottom") +
    labs(x = "log2count", y = "density", 
         title = "Make a ggplot like the students should")
```

# Run DESeq2

```{r}
# Creating DESeq2 dataset
col_data <- data.frame(condition = conditions)
rownames(col_data) <- colnames(counts)
dds <- DESeqDataSetFromMatrix(countData = counts, colData = col_data, design = ~ condition)

# Running DESeq2 analysis
dds <- DESeq(dds)

# Results
res <- results(dds)
head(res)


# Plotting the results
plotMA(res, main="DESeq2", ylim=c(-2,2))

# Histogram of p-values
hist(res$pvalue, breaks=50, col="skyblue", border="slateblue", main="Histogram of p-values")

```

# Conclusion

If nothing crashed and the plot above shows some blue dots around x = 20, y = -1.5 and triangles around x = 32, y = -2, we are good to go!
